Concurrent Programming in Harmony

### 2021-12-06 book/paper.tex
Fixed some "atom" leftovers

Harmony programs, and indeed the HVM,  manipulate Harmony values.
Harmony values are recursively defined:
they include booleans (\textbf{False} and \textbf{True}),
integers (but not floating point numbers),
strings (enclosed by double quotes),
sets of Harmony values,
and dictionaries
\index{dictionary}%
that map Harmony values to other Harmony values.

Harmony makes extensive use of dictionaries.
A dictionary maps keys to values.
Unlike Python, any Harmony value can be a key, including another
dictionary.
Dictionaries are written as
$\{ k_0: v_0, ~ k_1: v_1, ~ ... \}$.
If $d$ is a dictionary, and $k$ is a key, then the
following expression retrieves the Harmony value that $k$ maps to in $d$:
\begin{code}
$d$ $k$
\end{code}
The meaning of $d$ $a$ $b$ $...$ is ((($d$ $a$) $b$) $...$).
This notation is unfamiliar to Python programmers,
but in Harmony square brackets can be used
in the same way as parentheses, so you can express the same thing in the form
that is familiar to Python programmers:
\begin{code}
$d$[$k$]
\end{code}
However, if $d = \{ \mathtt{.count}: 3 \}$, then you can write
$d$.\texttt{count} (which has value 3) instead of having to write
$d$\texttt{[.count]} (although both will work).
Thus, using atoms, a dictionary can be made to look much like a Python object.

@@

It may seem strange that there is only one memory location.
However, this is not a limitation because Harmony values are unbounded trees.
The shared memory is a dictionary that maps atoms (names of shared
variables) to other Harmony values.
We call this a \emph{directory}.
\index{directory}%
Thus, a directory represents the state of a collection of variables named
by the atoms.
%
Because directories are Harmony values themselves,
directories can be organized into a tree.
Each node in a directory tree is then identified
by a sequence of Harmony values, like a path name in the file system
hierarchy.  We call such a sequence an \emph{address}.
\index{address}%
For example, in \autoref{fig:inc} the memory is a dictionary with two
entries: \texttt{.count} and \texttt{.done}.  And the value of entry
\texttt{.done} is a dictionary with keys 0 and 1.
So, for example, the address of \texttt{done[0]} is the sequence
[ \texttt{.done}, 0 ].
An address is itself a Harmony value.

@@

Tuples, lists, strings, and bags are all special cases of dictionaries.
Both tuples and lists map indexes (starting at 0) to Harmony values.
Their format is either \texttt{($e$, $e$, ..., $e$,)} or
\texttt{[$e$, $e$, ..., $e$,]}.
If the tuple or list has two or more elements, then the final comma
is optional.
A string is represented as a tuple of its characters.
Characters are one-character atoms, which can be expressed
in hexadecimal unicode using the syntax \texttt{.0xXX}.
A bag or multiset is a dictionary that maps a value to how many
times it occurs in the bag.

@@

\begin{itemize}
\item \texttt{False < True};
\item integers are ordered in the natural way;
\item atoms are lexicographically ordered;
\item program counters are ordered by their integer values;
\item dictionaries are first converted into a list of ordered (key, value)
pairs.  Then two dictionaries are lexicographically ordered by this
representation;
\item a set is first converted into an ordered list, then lexicographically
ordered;
\item an address is a list of atoms.  \texttt{None} is the empty list of atoms.
Addresses are lexicographically ordered accordingly;
\item contexts (\autoref{app:context}) are ordered first by name, then by program counter, then by the remaining content.
\end{itemize}

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
code & a list of HVM machine instructions \\
\hline
labels & a dictionary of atoms to program counters \\
\hline
variables & a dictionary mapping atoms to values \\
\hline
ctxbag & a bag of runnable contexts \\
\hline
stopbag & a bag of stopped contexts \\
\hline
choosing & if not \texttt{None}, indicates a context that is choosing \\
\hline
\end{tabular}
\vspace{1em}

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
name & the name of the main method that the thread is executing \\
\hline
argument & the argument given to the main method \\
\hline
program counter & an integer value pointing into the code \\
\hline
frame pointer & an integer value pointing into the stack \\
\hline
atomic & if non-zero, the thread is in atomic mode \\
\hline
readonly & if non-zero, the thread is in read-only mode \\
\hline
stack & a list of Harmony values \\
\hline
method variables & a dictionary mapping atoms (names of method variables) to values \\
\hline
thread-local variables & a dictionary mapping atoms (names of thread-local variables) to values \\
\hline
stopped & a boolean indicating if the context is stopped \\
\hline
failure & if not None, string that describes how the thread failed \\
\hline
\end{tabular}
\vspace{1em}

Details:
\begin{itemize}
\item The frame pointer points to the current \emph{stack frame},
which consists of the caller's frame pointer and variables, the argument to
the method, an ``invocation type'' (\texttt{normal}, \texttt{interrupt}, or
\texttt{thread}), and the return address (in case of \texttt{normal}).
\item A thread terminates when it reaches the \texttt{Return} instruction
of the top-level method (when the stack frame is of type \texttt{thread})
or when it hits an exception.  Exceptions include divide by zero,
reading a non-existent key in a dictionary, accessing a non-existent
variable, as well as when an assertion fails;
\item The execution of a thread in \emph{atomic mode} does not get interleaved
with that of other threads.
\item The execution of a thread in \emph{read-only mode} is not allowed
to update shared variables of spawn threads.
\item The register of a thread always contains a dictionary, mapping
atoms to arbitrary values.  The atoms correspond to the variable names
in a Harmony program.
\end{itemize}

>>>>

Harmony programs, and indeed the HVM,  manipulate Harmony values.
Harmony values are recursively defined:
they include booleans (\textbf{False} and \textbf{True}),
integers (but not floating point numbers),
strings (enclosed by single or double quotes),
sets of Harmony values, and dictionaries
\index{dictionary}%
that map Harmony values to other Harmony values.
Strings that start with a letter or an underscore and only contain
letters, digits, and underscores can be written without quotes by
preceding it with a dot.  So \texttt{.example} is the same string
as \texttt{"example"}.

Harmony makes extensive use of dictionaries.
A dictionary maps keys to values.
Unlike Python, any Harmony value can be a key, including another
dictionary.
Dictionaries are written as
$\{ k_0: v_0, ~ k_1: v_1, ~ ... \}$.
If $d$ is a dictionary, and $k$ is a key, then the
following expression retrieves the Harmony value that $k$ maps to in $d$:
\begin{code}
$d$ $k$
\end{code}
The meaning of $d$ $a$ $b$ $...$ is ((($d$ $a$) $b$) $...$).
This notation is unfamiliar to Python programmers,
but in Harmony square brackets can be used
in the same way as parentheses, so you can express the same thing in the form
that is familiar to Python programmers:
\begin{code}
$d$[$k$]
\end{code}
However, if $d = \{ \mathtt{.count}: 3 \}$, then you can write
$d$.\texttt{count} (which has value 3) instead of having to write
$d$\texttt{[.count]} or $d$\texttt{["count"]} (although any of those
will work).
Thus a dictionary can be made to look much like a Python object.

@@

It may seem strange that there is only one memory location.
However, this is not a limitation because Harmony values are unbounded trees.
The shared memory is a dictionary that maps strings (names of shared
variables) to other Harmony values.
We call this a \emph{directory}.
\index{directory}%
Thus, a directory represents the state of a collection of variables named
by the strings.
%
Because directories are Harmony values themselves,
directories can be organized into a tree.
Each node in a directory tree is then identified
by a sequence of Harmony values, like a path name in the file system
hierarchy.  We call such a sequence an \emph{address}.
\index{address}%
For example, in \autoref{fig:inc} the memory is a dictionary with two
entries: \texttt{.count} and \texttt{.done}.  And the value of entry
\texttt{.done} is a dictionary with keys 0 and 1.
So, for example, the address of \texttt{done[0]} is the sequence
[ \texttt{.done}, 0 ].
An address is itself a Harmony value.

@@

Tuples, lists, and bags are all special cases of dictionaries.
Both tuples and lists map indexes (starting at 0) to Harmony values.
Their format is either \texttt{($e$, $e$, ..., $e$,)} or
\texttt{[$e$, $e$, ..., $e$,]}.
If the tuple or list has two or more elements, then the final comma
is optional.
A bag or multiset is a dictionary that maps a value to how many
times it occurs in the bag.

@@

\begin{itemize}
\item \texttt{False < True};
\item integers are ordered in the natural way;
\item strings are lexicographically ordered;
\item program counters are ordered by their integer values;
\item dictionaries are first converted into a list of ordered (key, value)
pairs.  Then two dictionaries are lexicographically ordered by this
representation;
\item a set is first converted into an ordered list, then lexicographically
ordered;
\item an address is a list of Harmony values.  \texttt{None} is the empty list.
Addresses are lexicographically ordered accordingly;
\item contexts (\autoref{app:context}) are ordered first by name, then by program counter, then by the remaining content.
\end{itemize}

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
code & a list of HVM machine instructions \\
\hline
variables & a dictionary mapping strings to values \\
\hline
ctxbag & a bag of runnable contexts \\
\hline
stopbag & a bag of stopped contexts \\
\hline
choosing & if not \texttt{None}, indicates a context that is choosing \\
\hline
\end{tabular}
\vspace{1em}

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
name & the name of the main method that the thread is executing \\
\hline
argument & the argument given to the main method \\
\hline
program counter & an integer value pointing into the code \\
\hline
frame pointer & an integer value pointing into the stack \\
\hline
atomic & if non-zero, the thread is in atomic mode \\
\hline
readonly & if non-zero, the thread is in read-only mode \\
\hline
stack & a list of Harmony values \\
\hline
method variables & a dictionary mapping strings (names of method variables) to values \\
\hline
thread-local variables & a dictionary mapping strings (names of thread-local variables) to values \\
\hline
stopped & a boolean indicating if the context is stopped \\
\hline
failure & if not None, string that describes how the thread failed \\
\hline
\end{tabular}
\vspace{1em}

Details:
\begin{itemize}
\item The frame pointer points to the current \emph{stack frame},
which consists of the caller's frame pointer and variables, the argument to
the method, an ``invocation type'' (\texttt{normal}, \texttt{interrupt}, or
\texttt{thread}), and the return address (in case of \texttt{normal}).
\item A thread terminates when it reaches the \texttt{Return} instruction
of the top-level method (when the stack frame is of type \texttt{thread})
or when it hits an exception.  Exceptions include divide by zero,
reading a non-existent key in a dictionary, accessing a non-existent
variable, as well as when an assertion fails;
\item The execution of a thread in \emph{atomic mode} does not get interleaved
with that of other threads.
\item The execution of a thread in \emph{read-only mode} is not allowed
to update shared variables of spawn threads.
\item The register of a thread always contains a dictionary, mapping
strings to arbitrary values.  The strings correspond to the variable names
in a Harmony program.
\end{itemize}
